%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Unit 2: Data Management}
\label{chap:unit2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Total Time: 3.5 hours}


%================================================================
\section{Data Collection and Storage}
\label{sec:2.1}
%================================================================

\textbf{Time: 0.75 hours}

This lesson enables learners to design robust, ethical research plans and identify the optimal data collection method for high-quality, reproducible findings.

\subsection{Learning Objectives}

\begin{enumerate}
    \item Distinguish between qualitative and quantitative data.
    \item Apply methods for gathering, organizing, and cleaning data to ensure quality and integrity.
    \item Ensure ethical compliance of data collection, handling, and storage.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{itemize}
    \item From the Data Challenge, are the more than 100 data elements collected by the National Center for Health Statistics qualitative or quantitative?
    \item Were the SEER data organized and cleaned in a manner that led to your ability to reliably use them for secondary analyses? Why or why not?
    \item Why do you think they no longer report date and time of birth, and now only require week of birth?
\end{itemize}


%================================================================
\section{Metadata - Data About Data}
\label{sec:2.2}
%================================================================

\textbf{Time: 0.375 hours}

This lesson explores the essentials of metadata in biomedical research datasets, covering data collection methods, population, and context. Students will learn to identify high-quality metadata that supports reproducibility and distinguish it from inadequate metadata, ensuring robust and reliable research outcomes.

\subsection{Learning Objectives}

\begin{enumerate}
    \item Understand standard components of metadata on biomedical science research datasets: how the data were collected, on what population, under what circumstances, etc.
    \item Learn to distinguish between good and bad metadata for reproducibility.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{itemize}
    \item List 5 critical metadata categories that researchers need to know when reproducing findings.
    \item From the Mathew E. Hauer article called, ``Data Descriptors: Population projections for U.S. Counties by age, sex, and race,'' what metadata are provided about how the population data were collected?
\end{itemize}


%================================================================
\section{Data Representation}
\label{sec:2.3}
%================================================================

\textbf{Time: 0.375 hours}

This lesson examines how data can be represented in multiple ways, highlighting that each representation impacts task efficiency. Students will learn to select optimal data representations tailored to specific research tasks, balancing ease and complexity.

\subsection{Learning Objectives}

\begin{enumerate}
    \item Understand that the same data can be represented in many ways.
    \item Appreciate that each representation choice makes some tasks easier, but others more difficult.
    \item Learn how to choose a good representation for the task at hand.
\end{enumerate}

\subsection{Assessment Instrument}

The NCHS data are provided as a flat file with more than 100 variables. What is an alternative representation of these same data? Is the original flat file or your alternative schema more conducive to analyses, and why?

%================================================================
\section{Data Sharing}
\label{sec:2.4}
%================================================================

\textbf{Time: 2 hours}

\subsubsection{Data Sharing 101}

This lesson introduces the principles of Open Science, focusing on the NIH Data Management \& Sharing Policy's rationale and key components. Students will explore the FAIR Guiding Principles (Findable, Accessible, Interoperable, Reusable), learning their definitions and practical examples to promote transparent and reproducible research.

\textbf{Learning Objectives:}
\begin{itemize}
    \item Appreciate the foundations of Open Science
    \item NIH Data Management \& Sharing Policy: Rationale and Key components
    \item FAIR Guiding Principles: Definition and examples
\end{itemize}

\textbf{Assessment:} Define rationale behind NIH Data Management \& Sharing requirement; list 5 key components you should include in your 2-page Data Management Plan; list the 4 FAIR principles.

\subsubsection{Data Sharing - The Reality}

This lesson examines privacy and confidentiality concerns in Open Science and data sharing. Students will learn to differentiate between biomedical research types: bench science, human clinical trials, and animal models, and understand the unique data sharing implications for each, including ethical considerations and strategies to protect sensitive data while promoting transparency.

\textbf{Learning Objectives:}
\begin{enumerate}
    \item Learn about privacy/confidentiality concerns related to Open Science and data sharing.
    \item Articulate difference in types of biomedical research (bench science, human clinical trials, animal models) and what implications data sharing has for each.
\end{enumerate}

\textbf{Assessment:}
\begin{itemize}
    \item Create a 2-page Data Management and Sharing Plan following the NIH requirements for your analyses of the birthweight data challenge.
    \item Apply the FAIR principles (Findable, Accessible, Interoperable, Reusable) to your datasets.
\end{itemize}

% Required packages (add to your preamble if not already there):
% \usepackage{longtable}
% \usepackage{booktabs}
% \usepackage{ragged2e}
% \usepackage{array}
% \usepackage{pdflscape}

% \begin{landscape}

%     \section{Evaluation Rubric}

%     {\footnotesize

%     \begin{longtable}{%
%         >{\RaggedRight\arraybackslash}p{2.5cm}   % Component
%         >{\RaggedRight\arraybackslash}p{4.5cm}   % Excellent
%         >{\RaggedRight\arraybackslash}p{4.2cm}   % Good
%         >{\RaggedRight\arraybackslash}p{4.2cm}   % Satisfactory
%         >{\RaggedRight\arraybackslash}p{4.5cm}}  % Needs Improvement

%     \toprule
%     \textbf{Component} &
%     \textbf{Excellent (90--100\%)} &
%     \textbf{Good (80--89\%)} &
%     \textbf{Satisfactory (70--79\%)} &
%     \textbf{Needs Improvement ($<$70\%)} \\
%     \midrule
%     \endfirsthead

%     \toprule
%     \textbf{Component} &
%     \textbf{Excellent (90--100\%)} &
%     \textbf{Good (80--89\%)} &
%     \textbf{Satisfactory (70--79\%)} &
%     \textbf{Needs Improvement ($<$70\%)} \\
%     \midrule
%     \endhead

%     \midrule
%     \multicolumn{5}{r}{\textit{Continued on next page}} \\
%     \endfoot

%     \bottomrule
%     \endlastfoot

%     % Row 1
%     \textbf{1.\ Metadata -- Data About Data}
%     \newline\textbf{(20\%)}
%     &
%     Identifies 5 critical metadata categories with clear, relevant examples;
%     provides a thorough analysis of metadata from Hauer's article, including
%     how, on whom, and under what circumstances data were collected;
%     differentiates high-quality vs.\ poor metadata for reproducibility.
%     &
%     Lists most metadata categories with relevant examples; analysis from the
%     Hauer article is solid but lacks some detail; distinction between
%     good/bad metadata is mostly clear.
%     &
%     Lists some metadata categories but with limited examples or partial
%     relevance; Hauer's article analysis is basic; distinction is made at the
%     surface level.
%     &
%     Fewer than 5 categories or inaccurate examples; superficial or missing
%     analysis of Hauer article; shows little grasp of metadata quality for
%     reproducibility.
%     \\
%     \midrule

%     % Row 2
%     \textbf{2.\ Data Representation}
%     \newline\textbf{(20\%)}
%     &
%     Clearly describes alternative data representations and justifies which
%     schema best supports analysis tasks; insightfully compares the flat file
%     to the alternative, with thoughtful reasoning and examples; demonstrates
%     a strong understanding of how representation affects research.
%     &
%     Describes an alternative schema and makes a reasonable comparison to the
%     flat file, with some justification; demonstrates good understanding with
%     minor gaps.
%     &
%     Identifies a basic alternative, but justification/comparison to the flat
%     file is weak or incomplete; limited examples.
%     &
%     Incomplete, irrelevant, or unclear alternatives; little to no
%     justification or analysis of schema versus flat file; shows limited
%     understanding.
%     \\
%     \midrule

%     % Row 3
%     \textbf{3.\ Data Sharing 101: Open Science, NIH Policy, FAIR}
%     \newline\textbf{(30\%)}
%     &
%     Thoroughly defines rationale for NIH Data Management \& Sharing; lists 5
%     key components with well-explained relevance; accurately lists and
%     explains all 4 FAIR principles with practical examples for each.
%     &
%     Defines rationale and lists the key NIH components with basic relevance;
%     lists and explains all FAIR principles, with most examples being suitable.
%     &
%     Provides basic rationale and fewer than 5 NIH components; lists most
%     FAIR principles with limited or superficial examples.
%     &
%     Incomplete rationale and components, missing several critical points;
%     fails to list all FAIR principles or provides incorrect
%     explanations/examples.
%     \\
%     \midrule

%     % Row 4
%     \textbf{4.\ Data Sharing (The Reality): Privacy, Confidentiality,
%     Types of Biomedical Research}
%     \newline\textbf{(30\%)}
%     &
%     Creates a complete Data Management \& Sharing Plan tailored to the Texas
%     birthweight challenge, following NIH requirements; thoroughly applies
%     FAIR principles to the dataset; clearly articulates privacy concerns and
%     differences among bench, clinical, and animal model research, with
%     thoughtful analysis of sharing implications.
%     &
%     Plan meets most NIH requirements; applies FAIR principles with minor
%     gaps; reasonably discusses privacy and research type differences.
%     &
%     Plan basic or incomplete; applies FAIR principles at the surface level;
%     discusses privacy/research types with limited detail.
%     &
%     Missing/incomplete plan; fails to address FAIR principles, research
%     types, or privacy concerns meaningfully.
%     \\

%     \end{longtable}

%     } % end \footnotesize

% \end{landscape}


\begin{landscape}

    \section{Evaluation Rubric}

    {\footnotesize

    \begin{longtable}{%
        >{\RaggedRight\arraybackslash}p{2.5cm}   % Component
        >{\RaggedRight\arraybackslash}p{4.5cm}   % Excellent
        >{\RaggedRight\arraybackslash}p{4.2cm}   % Good
        >{\RaggedRight\arraybackslash}p{4.2cm}   % Satisfactory
        >{\RaggedRight\arraybackslash}p{4.5cm}}  % Needs Improvement

    \toprule
    \textbf{Component} &
    \textbf{Excellent (90--100\%)} &
    \textbf{Good (80--89\%)} &
    \textbf{Satisfactory (70--79\%)} &
    \textbf{Needs Improvement ($<$70\%)} \\
    \midrule
    \endfirsthead

    \toprule
    \textbf{Component} &
    \textbf{Excellent (90--100\%)} &
    \textbf{Good (80--89\%)} &
    \textbf{Satisfactory (70--79\%)} &
    \textbf{Needs Improvement ($<$70\%)} \\
    \midrule
    \endhead

    \midrule
    \multicolumn{5}{r}{\textit{Continued on next page}} \\
    \endfoot

    \bottomrule
    \endlastfoot

    % Row 1
    \textbf{1.\ Qualitative vs.\ Quantitative Data}
    \newline\textbf{(10\%)}
    &
    Accurately classifies all or nearly all of the 100+ NCHS data elements as
    qualitative or quantitative; provides clear, specific examples from the
    dataset to justify classifications; demonstrates nuanced understanding of
    the distinction and its implications for analysis and reproducibility.
    &
    Correctly classifies most data elements with reasonable examples;
    distinction between qualitative and quantitative is mostly clear and
    accurate, with minor gaps in justification or depth.
    &
    Provides a basic classification but with limited examples or incomplete
    reasoning; distinction is recognized but not consistently or accurately
    applied to the NCHS dataset.
    &
    Classification is missing, incorrect, or conflates qualitative and
    quantitative data; examples absent or irrelevant; shows little
    understanding of the distinction.
    \\
    \midrule

    % Row 2
    \textbf{2.\ Data Organization, Cleaning, \& Quality}
    \newline\textbf{(10\%)}
    &
    Provides a thorough, well-reasoned evaluation of whether the SEER data
    were organized and cleaned for reliable secondary use; identifies specific
    strengths and weaknesses in data organization; connects data quality
    decisions to reproducibility and research integrity.
    &
    Offers a reasonable evaluation of SEER data organization and cleaning;
    identifies most relevant quality issues with adequate justification;
    connection to reproducibility is mostly developed, with minor gaps.
    &
    Provides a basic evaluation but analysis is superficial or one-sided;
    acknowledges data quality issues without substantive discussion of their
    implications for secondary analysis.
    &
    Evaluation missing or incorrect; fails to engage meaningfully with SEER
    data organization or cleaning; no connection drawn to data quality,
    reliability, or reproducibility.
    \\
    \midrule

    % Row 3
    \textbf{3.\ Ethical Compliance in Data Collection \& Storage}
    \newline\textbf{(10\%)}
    &
    Provides a well-reasoned, nuanced explanation for why date and time of
    birth are no longer reported; situates the decision within ethical,
    privacy, and policy contexts; demonstrates sophisticated understanding of
    how ethical compliance shapes data collection and storage decisions.
    &
    Offers a reasonable explanation with some ethical and privacy grounding;
    demonstrates solid understanding of how compliance considerations affect
    data collection practices, with minor gaps.
    &
    Provides a basic explanation and acknowledges ethical or privacy
    motivations but lacks depth; recognizes compliance as a factor without
    fully analyzing its implications.
    &
    Explanation missing or purely speculative; no grounding in ethical,
    privacy, or policy context; shows little understanding of how compliance
    shapes data collection and storage decisions.
    \\
    \midrule

    % Row 4
    \textbf{4.\ Metadata -- Data About Data}
    \newline\textbf{(15\%)}
    &
    Identifies 5 critical metadata categories with clear, relevant examples;
    provides a thorough analysis of metadata from Hauer's article, including
    how, on whom, and under what circumstances data were collected;
    differentiates high-quality vs.\ poor metadata for reproducibility.
    &
    Lists most metadata categories with relevant examples; analysis from the
    Hauer article is solid but lacks some detail; distinction between
    good/bad metadata is mostly clear.
    &
    Lists some metadata categories but with limited examples or partial
    relevance; Hauer's article analysis is basic; distinction is made at the
    surface level.
    &
    Fewer than 5 categories or inaccurate examples; superficial or missing
    analysis of Hauer article; shows little grasp of metadata quality for
    reproducibility.
    \\
    \midrule

    % Row 5
    \textbf{5.\ Data Representation}
    \newline\textbf{(15\%)}
    &
    Clearly describes alternative data representations and justifies which
    schema best supports analysis tasks; insightfully compares the flat file
    to the alternative, with thoughtful reasoning and examples; demonstrates
    a strong understanding of how representation affects research.
    &
    Describes an alternative schema and makes a reasonable comparison to the
    flat file, with some justification; demonstrates good understanding with
    minor gaps.
    &
    Identifies a basic alternative, but justification/comparison to the flat
    file is weak or incomplete; limited examples.
    &
    Incomplete, irrelevant, or unclear alternatives; little to no
    justification or analysis of schema versus flat file; shows limited
    understanding.
    \\
    \midrule

    % Row 6
    \textbf{6.\ Data Sharing 101: Open Science, NIH Policy, FAIR}
    \newline\textbf{(25\%)}
    &
    Thoroughly defines rationale for NIH Data Management \& Sharing; lists 5
    key components with well-explained relevance; accurately lists and
    explains all 4 FAIR principles with practical examples for each.
    &
    Defines rationale and lists the key NIH components with basic relevance;
    lists and explains all FAIR principles, with most examples being suitable.
    &
    Provides basic rationale and fewer than 5 NIH components; lists most
    FAIR principles with limited or superficial examples.
    &
    Incomplete rationale and components, missing several critical points;
    fails to list all FAIR principles or provides incorrect
    explanations/examples.
    \\
    \midrule

    % Row 7
    \textbf{7.\ Data Sharing (The Reality): Privacy, Confidentiality,
    Types of Biomedical Research}
    \newline\textbf{(15\%)}
    &
    Creates a complete Data Management \& Sharing Plan tailored to the Texas
    birthweight challenge, following NIH requirements; thoroughly applies
    FAIR principles to the dataset; clearly articulates privacy concerns and
    differences among bench, clinical, and animal model research, with
    thoughtful analysis of sharing implications.
    &
    Plan meets most NIH requirements; applies FAIR principles with minor
    gaps; reasonably discusses privacy and research type differences.
    &
    Plan basic or incomplete; applies FAIR principles at the surface level;
    discusses privacy/research types with limited detail.
    &
    Missing/incomplete plan; fails to address FAIR principles, research
    types, or privacy concerns meaningfully.
    \\

    \end{longtable}

    } % end \footnotesize

\end{landscape}