---
title: "Create a new project"
author: "SW"
format: html
editor: visual
---

This file is meant to complete the 1st step in the workflow for a new consultanting case and 2nd step move the template to the new R project

1st step: When a new case got assigned, a new R project will be created in the consultant specified directiory. SW habit: Casenumber_Clienlastname_clientfirstname.

```{r}
#| label: load-pkg
library(usethis)
## file folder setup
library(fs)   # creat folder dir_create()
library(here) # easy to do cross reference in the project.


```

```{r}
#| label: create new proj

# I wish to use usethis pkg create_project(), however it won't do without git. Until we find a solution, we will manually delete git files.

#One reason I like use this code to create project is the speed. A lot of time it takes a while to create a new project and wait for it open. With the code, it takes seconds and open automatically.

create_project(path = "P:/Consulting/Secure/THAA/C1340 Suneja (NK)",open = TRUE, rstudio = TRUE)



```

2nd step: copy the setup file.

```{r}
#| label: move set up file

file_copy("C:/Users/wangsic1/OneDrive - Michigan State University/CSTAT main/Template/2nd_Set up folders_8_7_2023.qmd", "P:/Consulting/Secure/THAA/C1340 Suneja(NK)", overwrite = FALSE)

```

Once the setup file created in the new R project, we will open set up file to create folders.

SW: Thinking combine this with set up.
