# 1st step: github.com to create a new repo from template, *please do this first*

Copy the HTTPS.

# 2nd: When a new case got assigned, a new R project will be created in the consultant specified directiory. SW habit: Casenumber_Clienlastname_clientfirstname. From git

```{r}
#| label: load-pkg
library(usethis)
## file folder setup
library(fs)   # creat folder dir_create()
library(here) # easy to do cross reference in the project.


```

Insert the git HTTPS below to pull the template setting and start R project

```{r}
usethis::create_from_github(
  "https://github.com/CSTAT-CVM/Testing.git",
  destdir = "C:/Users/wangsic1/OneDrive - Michigan State University/CSTAT main/2023 Summer/Case/"
)
```

# Since I put the setup files in git template, I do not need the 3rd step here.

# BTW, I should keep my setup files updated on git.

3rd step: copy the setup file.

```{r}
#| label: move set up file

file_copy("C:/Users/wangsic1/OneDrive - Michigan State University/CSTAT main/Template/2nd_Set up folders_8_7_2023.qmd", "C:/Users/wangsic1/OneDrive - Michigan State University/CSTAT main/2023 Summer/Case/C5858_Wang_Sichao", overwrite = FALSE)

```
