---
title: "C1000: Short title"
author: 
  - name: Name
    email: email@msu.edu
    degrees: PhD
    orcid: 0000-0003-0931-7883
    affiliations: 
      - name: Michigan State University, Center for Statistical Training and Consulting
date: "`r format(Sys.time(), '%Y-%m-%d, %H:%M:%S %Z')`"
title-block-banner: "#01665e"
format: 
  html: 
    toc: true 
    toc-depth: 4
    toc-location: left 
    toc-title: Table of contents
    number-sections: true
    number-depth: 4
    code-fold: false 
    code-tools: false 
    embed-resources: true 
    theme: default 
css: idastyle.css
execute: 
  eval: true 
  echo: false #fenced
  output: true 
  message: false 
  warning: false 
  error: false 
  include: true 
knitr:
  opts_chunk:
    out.width: "150%"
bibliography: references.bib
editor:
  markdown:
    wrap: sentence
---


# Project Overview:

**Title: Full title**

**Project PI's:**

<p>

-   Name1 (email1\@msu.edu)
-   Name2 (email2\@msu.edu)

</p>

<h3>Data sets</h3>



</p>
 
<h3>Primary Objective:</h3>
 
<p>**The objective of this study is ...**</p>

<h3>Secondary Objective:</h3>
 
<p>
 
1.  **Objective 1**
2.  ~~Objective 2~~ [Variable `var` is missing]{style="color:red;"}


</p>
 
<h3>References:</h3>
 
Initial data analysis: @tsr2022, @idacheck2024 and @idacheck2024B.
 
 
::: {.callout-note title="Project Updates" collapse="false"}
-   Date 1: Per email from … decision
-   Date 2: Per meeting … decision
:::


Updates detail on DATE:

-   [Update:]{style="color:green;"} In @sec-idamissing something was added.


```{r}
#| label: load-packages
#| include: false

pacman::p_load(## file folder setup
               fs,   # creat folder dir_create()
               here, # easy to do cross reference in the project.
               
               ## Data step
               readr,
               readxl, # read_excel()
               tidyverse,
               janitor,
               stringr,
               
               ## IDA
               Hmisc,
               naniar, ## for missing data 
               VIM, ## for missing data 
               
               ## Model
               meta,
               metafor,
               dmetar, #for find.outliers()
               lme4,
               glmmTMB,
               broom,
               emmeans,
               survival,
               performance,
               
               ##repeatibility
               irr,# for agreement
               psych,# for agreement
               confintr,#for ci_cor()
               DescTools,
               
               
               ## result reporting
               ### graph
               ggplot2,
               survminer,
               gridExtra,
               ggpubr,# for cowplot and plotgrid
               ### table
               gt,
               gtsummary,
               knitr,        # for kable()
               sjPlot,      # tab_model()
               gtExtras,# for get table theme like gt_theme_538()
               patchwork
               )


set.seed(42004063)



###################### color scales

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

colshue = gg_color_hue(10)

#plot(1:10, pch = 16, cex = 2, col = colshue)

library(scales)
hex <- hue_pal()(6)


###### colorblind colors

# colorbrewer
colsbrewer=c("#8c510a", "#d8b365", "#5ab4ac", "#01665e")

# Okabe-Ito
cols10<-c("#E69F00","#0072B2","#D55E00","#009E73","#F5C710","#56B4E9","#CC79A7","#999999","#000000","#01665e")



#https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html

```

```


```{r}
#| label: setup-folder
#| eval: false
 
# Create the default folder/subfolder. Only needs to be done once.
 
dir_create(here("Data"))  # store raw data, clean data.
dir_create(here("Data/Source")) 
dir_create(here("Data/Clean"))
 
dir_create(here("Documents"))  # all the docs
dir_create(here("Code"))       # r code
dir_create(here("Emails"))    # correspondence
 
# Declare location of this script relative to the project root directory.
here::i_am(path = "Code/idatemplate.qmd")  #save the qmd under code first and then run this line.
 
# dir_ls()    # briefly check
# dir_tree()  # Show the tree, not necessary, takes time.
# dir_tree("Data")
 
```
 
```{r}
#| label: import-data
#| eval: false
 
# This section is for reading the raw data, do some preprocessing steps (e.g. new variable definitions).
# Then cleaned data will be saved as rds.
# Only needs to be done once.




# Example of preprocessing

### 1st Step ###---: import dataset and clean column names.

df0 <- read_excel(here("Data/Raw/datname.xlsx"), 
    sheet = "Sheet1",na = c("","NA")) %>% 
  clean_names() %>% 
  remove_empty(which = c("rows", "cols"), cutoff = 1, quiet = TRUE) %>% 
  rename(comorb.other=other,readmit_date=readmit) 

# base funtion make.names(casefold(names(df)), allow_=FALSE) works the same as clean_names to make names lower case, advantage is less dependency.


### 2nd step ###---: examine variables

# str(df) #check the nature of variables # I am happy with the date variables, numeric variables. character variables need to convert to factors.
# describe(df) # check levels of character vars


### 3rd Step ###---: recode of variables

df1 <- df0 |>     
  # first change x to 1 to indicate yes
  mutate(across(.cols = c("htn" , "dm","cardiac", "copd"),
                .fns = ~case_when(.x == "x" ~ "1",
                       TRUE ~"0"))) |>   # change  NA to 0
  mutate(variable1 =case_when(variable1=="Yes"~"yes",
                   is.na(variable1)~"no",
                          TRUE ~variable1), # change values
  variable2 = tolower(variable2)) |> 
  mutate_if(is.character, tolower) |> # to remove inconsistency due to upper/lower cases
  mutate_if(is.character, as.factor) 

### 4th Step ###---: define new variables. 

df2 <- df1 |>
  mutate(var1_numeric=as.numeric(as.character(var1)), 
         var1_factor=factor(ifelse(as.character(var1) != "lp","not_lp","lp"),level=c("not_lp","lp")), # make sure preserve NA sum(is.na(df2$var1_factor))
         number_var2_numeric=as.numeric(as.character(var2)),
         number_var2_factor=factor(ifelse(as.character(var2)!="lp","not_lp","lp"),level=c("not_lp","lp")),
         number_var2=fct_relevel(var2,"lp", "1","2","≥3"),
         var2=fct_relevel(number_abscesses,"lp", "1","2","3"),
         t2readmit=as.numeric(difftime(readmit_date,date_of_discharge,units="day")),
         readmit=factor(ifelse(is.na(readmit_date),"no","yes"))),
  mutate(a_vs_b=ifelse(a=="val1","a","b")) |> 
  mutate(dc_abx=case_when(
     dc_abx=="med1" & oral_iv=="po"~ "med1",
     dc_abx=="med2" & oral_iv=="po" ~ "med2",
     is.na(dc_abx) ~ NA,
     .default = "other"
    )) |> 
    
# Check results of this code and make comments here 
  
# e.g. examine cross tables of derived and original categorical variables
# e.g. do calculated time differences correspond to variables in dataset?
  
  
### 5th Step ###---: deselect variables that are not needed for analysis 

df_clean <- df2 |> select(-mrn,-name)


## Save dataset (possible multiple datasets) after data preparation. 
# Only needs to be done once.
 
saveRDS(df_clean, file = here("Data/Clean/datname.rds"))

```
 

 
 
```{r}
#| label: ida-load-cleandata
#| include: false
 
df_ida <- readRDS(here("Data/Clean/datname.rds"))
 
 
# str(df_ida)

```
 
 
```{r}
#| label: ida-variables-grouping
#| include: false


 
# This section meant to group variables according to their purpose in the analysis or defined by client.
 
demog_vars<-c()
lab_vars<-c()
explan_vars<-c()
outcome_vars<-c()
time2event_vars<-c()
date_vars<-c()
 
```





# Initial Data Analysis (IDA)

The aim of IDA is to examine data properties to ensure transparency and integrity of preconditions to conduct appropriate statistical analyses and correct interpretation of the results to answer pre-defined research questions. 

**Some Notes about data:** 

-   Note 1
-   Note 2



 
## IDA: Participation profile {#sec-participation}
 
* Provide number of time points and intervals at which measurements are taken, using the time metric that best reflects the time of inclusion in the study. 

* Provide the number of participants who attended the assessment by the time metric(s) for the analysis.

### Dates

* Do date checks for the time frame in the study protocol

* Describe the time metric and corresponding time points specified in the analysis strategy. 

#### Admission Dates

```{r }
#| label: fig-plot-admit-dates
#| fig-cap: "Monthly Counts of admission based on admission date"


caption_text <- paste0("Each bar is one month." )

df_ida |> 
  dplyr::mutate(or.month = floor_date(ymd_hms(admitdt), 'month')) |> 
  janitor::tabyl(or.month) |> 
  as.data.frame() |> 
  ggplot(aes(x=ymd(or.month), y=n)) +
  geom_bar(stat="identity",fill="lightblue")+ 
  # geom_vline(xintercept = as.numeric(ymd("2022-9-01")),linetype=4, colour="black")+ #the last cohort 1 is 2022-9-1
  xlab("Admission Dates") +
    ylab("Count") + 
# scale_x_date(date_breaks = "1 year", date_labels =  "%Y") + 
    theme_bw() +
  ggtitle("Counts of Addmission Dates") + 
   labs(caption = caption_text)+
    theme(axis.text.x=element_text(angle=60, hjust=1),
        plot.title = element_text(face = "bold",  hjust = 0.5),
        plot.caption = element_text(face = "bold"  )) 

```

@fig-plot-admit-dates shows the first data point is in `r floor_date(ymd_hms(min(df_ida$admitdt)), 'day')` and last data point is in `r floor_date(ymd_hms(max(df_ida$admitdt)), 'day')`.






## IDA: Missing values {#sec-idamissing}
 
Provide the number and proportion of missing values for each variable at each time point as appropriate for fixed or time-varying variables. Describe missingness stratifying the summaries by variables that might influence the frequency of missing values, if relevant (for example: structural variables or levels of measurement).

 
### Prevalence of missing values 

#### For each variable or groups of variables

Provide the number and proportion of missing values for each variable at each time point as appropriate for fixed or time-varying variables. Describe missingness stratifying the summaries by variables that might influence the frequency of missing values, if relevant (for example: structural variables or levels of measurement).


```{r}
#| label: tbl-ida-missingness-by-variable-demog_vars
#| tbl-cap: "Summary of missing values for demographic variables"


df_ida |>
  select(demog_vars) |>
  miss_var_summary() |>
  gt() |>
    gt::cols_label(
    variable = "Variable",
    n_miss = "Missing (count)",
    pct_miss = "Missing (%)"
  ) |>
  fmt_number(
    columns = vars(pct_miss),
    decimals = 1
  ) |>
  tab_header(title = "Summary of missing values for demographic variables ") |>
  gt_theme_538()

```

@tbl-ida-missingness-by-variable-demog_vars includes the baseline characteristics variables of the patients.

#### Complete cases

Describe number of complete observations when considering outcome and explanatory variables for any model.
 
```{r}
#| label: tbl-ida-missingness-complete-case
#| tbl-cap: "Summary of complete cases by sets of variables"


# Number of available complete cases (outcome and predictors):
# from Regression without regrets github code

 ard_all <- df_ida |> 
    #filter(KEY_PRED_FL01 == "Y") |>
    # select(SUBJID, PARAMCD, AVAL, SEXC, AGEGR01, AGEGR01C) |>
    # pivot_wider(names_from = PARAMCD, values_from = AVAL, values_fill = NA) |>
    # select(-SUBJID) |>
    # group_by(SEXC, AGEGR01, AGEGR01C) |>
    miss_case_table() |> 
    filter(n_miss_in_case == 0) |>
    mutate(Set = "All variables in the dataset", 
           row_order = 1) |>
    ungroup()

ard_demog <- df_ida |> 
    #filter(KEY_PRED_FL01 == "Y") |>
    select(demog_vars) |>
    # pivot_wider(names_from = PARAMCD, values_from = AVAL, values_fill = NA) |>
    # select(-SUBJID) |>
    # group_by(SEXC, AGEGR01, AGEGR01C) |>
    miss_case_table() |> 
    filter(n_miss_in_case == 0) |>
    mutate(Set = "Demographic variables", 
           row_order = 2) |>
    ungroup()


final_ard |> 
  select(Set, n_cases, pct_cases ) |>
  gt() |>
   cols_label(
    Set  = "Set",
    n_cases = "Complete (count)",
    pct_cases = "Complete (%)"
  ) |>
  fmt_number(
    columns = c(pct_cases),
    decimals = 1
  ) |>
  tab_header(title = "Summary of complete cases by sets of variables") |>
  gt_theme_538() 

```


### Missing values by Case

```{r}
#| label: fig-missing-data-using-vis_miss
#| fig-cap: "Missing data patten"
#| warning: false


# filter
df_ida |> 
  select(where(function(x) sum(is.na(x)) / length(x) > 0))|> # only show the ones with missing values
  vis_miss()

```

@fig-missing-data-using-vis_miss shows patterns of missingness for variables with at least has one missing value.


## IDA: Univariate descriptions {#sec-idaunivariate}
 
This section summarizes each variable by using appropriate summary statistics to make sure data properties are as expected and support the choices of  statistical methods.

*Optional: If this is a longitudinal study, describe variables at each time point.*
 
### Demographic variables

Summarize the demographic variables with numerical and graphical summaries.

```{r }
#| label: ida-describe-demog_vars


d <- df_ida |> 
  dplyr::select(any_of(demog_vars)) |> 
 Hmisc::describe()

Hmisc::html(d)
# Hmisc::html(d, size=80, scroll=TRUE)

```


### Explanatory variables

Summarize the explanatory variables with numerical and graphical summaries at baseline.

```{r}
#| label: fig-highresolution-hist
#| fig-cap: "High definition of histograms for variable"

df_ida |> 
  ggplot(aes(x=varname))+
  geom_histogram(nbin=100, color="grey")+
    xlab("varname") +
    ylab("Count") + 
    theme_bw() 

```

### Outcome variables

Summarize the outcome variable  with numerical and graphical summaries at baseline.

 


## IDA: Multivariate descriptions {#sec-idamultivariate}


### Association

Visualize and summarize the association of each predictor with the structural variables. 


```{r }
#| label: tbl-ida-multi-compare-demog_vars
#| tbl-cap: "Demographic variables by var"




formula_multi <- formula(paste0(paste0(demog_vars, collapse = " + "),"~ structuralvar"))

s <-  Hmisc::summaryM(formula_multi, data=df_ida, overall=TRUE, test = TRUE )


Hmisc::html(
  s,
  caption = 'Demographic variables by structural variable',
  exclude1 = TRUE,
  npct = 'both',
  digits = 3,
  pctdig =3,
  pdig = 3, eps = 0.001, #pvalue digits
  prmsd = TRUE,
  brmsd = TRUE,
  Nsize=TRUE
)

```

@tbl-ida-multi-compare-demog_vars shows XYZ. 


### Correlation

Quantify pairwise correlations between all key predictors 

### Interactions, if applicable

Evaluate bivariate distributions of the predictors specified in interactions. Include appropriate graphical displays.

### Clustering, optional

Visualize clustering of predictors using a dendrogram to show closely associated predictors 




## IDA longitudinal aspects {#sec-idalongitudinal}

### Profiles

Summarize changes and variability of variables within subjects, e.g. profile plots (spaghetti-plots) for groups of individuals.

### Trends

* Describe numerically or graphically longitudinal(average) trends of the outcome variable.

* Describe numerically or graphically the longitudinal trends of the time-varying variables.

### Time-to-event variables

Visualize with cumulative incidence plots


 


# Statistical analyses
 
In this section, we will follow the analysis strategy outlined in the SAP for each outcome.
 
## Model(s) for primary outcome


## Model(s) for secondary outcome




# Session Info

```{r}
#| include: false


sessionInfo()

```




# References



